---
title: "Comprehensive data analysis report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive data analysis report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document contains the entire data analysis pipeline 
aphantasiaCognitiveClustering was built for and the associated results reported
in the article "*Unsupervised clustering reveals spatial 
and verbal cognitive profiles in aphantasia and typical imagery*" 
(Delem et al., 2025). 

The main dataset used for the analysis is included in the package as the
`study_data` data frame.

```{r setup, message = FALSE}
library(aphantasiaCognitiveClustering)

head(study_data) |> knitr::kable()
```

# VVIQ group analysis

We first analysed the data in light of the VVIQ groups, examining differences 
between individuals with aphantasia and controls.

## Demographic variables

We tested the education level, fields of study and occupations of the
participants for any association with the two groups. The 
`get_association_models()` function was designed to compute Bayes Factors for 
associations between a grouping variable (here `group`) and the variables
`education`, `field` and `occupation` in the data. The computation relies on the
[BayesFactor](https://cran.r-project.org/package=BayesFactor) package.

The associated function `format_association_table()` extracts the contingency
table from the nested output of `get_association_models()` and formats it
into a clean table.

```{r demographic_sassociations}
df_association <- get_association_models(study_data, group)

df_association |> dplyr::select(Variable, log_bf10)

format_association_table(df_association, "Education")

format_association_table(df_association, "Field")

format_association_table(df_association, "Occupation")
```

## Main score variables

We analysed the main questionnaire and task score variables in the data, 
which are: the VVIQ, the three OSIVQ subscales (object, spatial, verbal), the
seven Psi-Q subscales (vision, auditory, smell, taste, touch, sensations, 
feelings), the Raven matrices, the SRI (spatial test), the digit span, the
spatial span, the similarities test, the WCST (executive functions test), and 
the reading comprehension task.

We fitted Bayesian linear models to the various variables to model them 
with participant groups as categorical predictors and age as a continuous 
covariate so as to control for the potential influence of the latter. We used
the [bayestestR](https://cran.r-project.org/package=bayestestR) package to 
compute the Bayes Factor for inclusion of the grouping variable and the age 
covariate in the models. To get Bayes Factors for the contrasts between the 
groups, we fitted models for the prior and posterior distributions using the
[rstanarm](https://cran.r-project.org/package=rstanarm) package (with default
settings) then used the [emmeans](https://cran.r-project.org/package=emmeans) 
package to compute the contrasts.

The `get_mean_sd()` function provides a quick way to compute and format the
mean and standard deviation of the variables for each group, while the
`get_bf_inclusion()` and `get_contrast_bf()` functions take care of the two 
modelling steps described above. All three are wrapped in the 
`get_full_model_table()` function, which is designed to run the full analysis 
pipeline for each variable in any long format data frame with the appropriate 
columns. The `get_longer()` function is designed to perform this transformation
from wide to long format easily, returning clean column names and values.

```{r get_longer_example}
study_data |> 
  get_longer() |> 
  dplyr::select(id, Group, Variable, value) |> 
  head()
```

The code to fit everything is straightforward:

```{r get_full_model_table_groups, eval = FALSE}
study_data |> get_longer() |> get_full_model_table(Group)
```

The number of models to fit is quite large, so this code does not run here. 
However, the results are saved natively in the package in the `models_list`
object. The resulting data frame for the two groups is the first item of this
list:

```{r models_list_groups}
models_list$group_models |> 
  filter_study_variables("original") |> 
  dplyr::select(!Comparison) |> 
  knitr::kable()
```

The `filter_study_variables()` function is a convenience function to filter
a set of variables from our study. Here the table contains only the original
raw scores, leaving out the transformed variables created later in the analyses.

Two functions are provided to visualise the data, 
`plot_score_violins()` (used in the paper for the groups) and 
`plot_score_radars()` (used later in the paper for the clusters). The latter
requires a `Cluster` column (I know, not a great design), so we'll use it later. 
`plot_score_violins()` uses the [see](https://cran.r-project.org/package=see) 
package to plot half-violins, while `plot_score_radars()` uses the
[superb](https://cran.r-project.org/package=superb) package to plot radar 
charts. The `scale_vars()` function is used to rescale variables between 0 and 1 
to plot them easily on the same scale.

```{r}
#| label: plot_score_violins_groups
#| fig.width: 7
#| fig.height: 3.5
#| fig.alt: > 
#|  Violin plots of the original score variables by VVIQ group.

study_data |> 
  scale_vars() |> 
  get_longer() |>
  filter_study_variables("original") |> 
  plot_score_violins()
```

# Cluster analysis

Faced with the lack of explanatory power of the VVIQ groups on many variables,
we set out to explore the structure of the data using clustering analysis.
This first required to study the relationships between the variables and
select the most relevant to describe the sample.

## Correlations

We used partial correlations to identify the strongest links between variables 
for variable reduction while reducing the bias from potential spurious 
correlations. The `correlate_vars()` function wraps the function from the
[correlation](https://cran.r-project.org/package=correlation) package to compute
these correlations and conveniently renames the variables in the resulting
object. Two plotting functions are provided to visualise the results, 
`plot_score_cor_matrix()` and `plot_score_cor_graph()`. The two are wrapped in
`plot_score_cor_joint()` to produce a joint plot of the correlation matrix and
the graph visualisation of the correlations with a specific overlapped layout
using the [patchwork](https://cran.r-project.org/package=patchwork) package.

```{r}
#| label: correlations
#| fig.width: 7
#| fig.height: 5
#| fig.alt: > 
#|  Joint plot of the correlation matrix and graph of the partial correlations
#|  between the study variables.

study_data |> correlate_vars(partial = TRUE) |> plot_score_cor_joint()
```

As represented by the coloured nodes, the partial correlations allowed to
identify related variables that could be relevant to merge and keep for the
clustering analysis. The clustering process was conducted on the seven 
following variables: *Visual imagery* (VVIQ + OSIVQ-O + Psi-Q V), *Sensory
imagery* (Psi-Q Sm+Ta+To+Se+Fe), *Spatial imagery* (OSIVQ-S + SRI), 
*Verbal strategies* (OSIVQ-V), *Raven + Digit span*, *Verbal reasoning* 
(Similarity) and *Spatial span*.  

## Number of clusters

We chose to use a model-based method for clustering (Gaussian Mixture 
Modelling) with the [mclust](https://cran.r-project.org/package=mclust) package.
The `cluster_selected_vars()` function computes the clustering on the 
variables described above. Internally, it calls the `scale_reduce_vars()`
function, which creates all the reduced composite variables identified with the
partial correlations above and rescales them between 0 and 1.

The optimal model and number of clusters was determined using the Bayesian
Information Criterion (BIC) implemented in the mclust package. The 
`plot_clusters_bic` uses 
[factoextra](https://cran.r-project.org/package=factoextra) to represent the
results of this model comparison visually.

```{r}
#| label: plot_bic
#| fig.width: 5
#| fig.height: 4.5
#| fig.alt: > 
#|  Plot of the BIC values for different clustering models and numbers of
#|  clusters.

study_data |> cluster_selected_vars() |> plot_clusters_bic()
```

## Clustering results

The `merge_clusters()` function allows to merge the original data, the reduced
variables and the cluster results easily.

```{r merge_clusters}
df_merged <- merge_clusters(
  df_raw     = study_data,
  df_red     = scale_reduce_vars(study_data),
  clustering = cluster_selected_vars(study_data)
)
colnames(df_merged)

df_merged |> 
  dplyr::reframe(
    .by = c(group, cluster),
    n       = dplyr::n(),
    vviq    = mean(vviq),
    object  = mean(osivq_o),
    spatial = mean(osivq_s),
    verbal  = mean(osivq_v)
  ) |> 
  dplyr::arrange(cluster)
```

There are two clusters containing only aphantasic (cluster A) or control 
(cluster C) participants, while another cluster (cluster B) is mixed. By 
dividing the latter into two "sub-clusters", we created a new `subcluster`
variable to analyse. The radar charts are a convenient way to visualise the
scores of the clusters and sub-clusters on the reduced variables:

```{r}
#| label: plot_score_radars_clusters
#| fig.width: 7
#| fig.height: 3.5
#| fig.alt: > 
#|  Radar charts of the clusters (left) and subclusters (right) on the reduced
#|  variables.

# Attaching superb is necessary to avoid bugs with the 1.0
library(superb)

df_merged_long_reduced <- 
  df_merged |> 
  scale_vars() |> 
  get_longer() |> 
  filter_study_variables("reduced")

plot_score_radars(df_merged_long_reduced, Cluster, r_off = 6, l_off = 6) + 
  plot_score_radars(df_merged_long_reduced, Subcluster, r_off = 6, l_off = 6)
```

We computed the Bayes Factors for association of the clusters and subclusters
with the demographic variables:

```{r demographic_associations_clusters}
df_asso_clusters <- get_association_models(df_merged, cluster)
df_asso_subclusters <- get_association_models(df_merged, subcluster)

df_asso_clusters |> dplyr::select(Variable, log_bf10)
df_asso_subclusters |> dplyr::select(Variable, log_bf10)

df_asso_clusters |> format_association_table("Education")
df_asso_clusters |> format_association_table("Field")
df_asso_clusters |> format_association_table("Occupation")

df_asso_subclusters |> format_association_table("Education")
df_asso_subclusters |> format_association_table("Field")
df_asso_subclusters |> format_association_table("Occupation")
```


Just like the group models, the code to model all the score variables with the 
clusters and sub-clusters is straightforward:

```{r get_full_model_table_clusters, eval = FALSE}
df_merged_long <- 
  df_merged |> 
  scale_vars() |> 
  get_longer()

cluster_models    <- get_full_model_table(df_merged_long, Cluster)
subcluster_models <- get_full_model_table(df_merged_long, Subcluster)
```

... And likewise, it takes a while to run, so models have been pre-computed
and the results stored in the `models_list` object.

Here are the Bayes Factors for inclusion of the predictors in the models for
each variable, for the clusters and subclusters:

```{r models_list_clusters_bfi}
models_list$cluster_models |> 
  dplyr::select(1:7) |> 
  dplyr::distinct() |> 
  knitr::kable()
models_list$subcluster_models |> 
  dplyr::select(1:7) |> 
  dplyr::distinct() |> 
  knitr::kable()
```

Finally, we can get the pairwise comparisons between the clusters and between
the subclusters, with the Bayes Factors for the contrasts:

```{r models_list_clusters_contrasts}
models_list$cluster_models |> 
  dplyr::select(1, 8:11) |> 
  knitr::kable()
models_list$subcluster_models |> 
  dplyr::select(1, 8:11) |> 
  knitr::kable()
```

------

...And this concludes this data analysis report.

```{r session_info}
sessioninfo::session_info()
```

